---
title: "Comparison of health and economic damages from storms and weather events in the United States"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE)
library(plyr)
library(ggplot2)
library(reshape2)
```

## Abstract

The main objective is to compare how different the storm and weather events are considering health and economic perpesctives.
We will analyse data from the National Oceanic & Atmospheric Administration. The data was released the 08/03/2007 and is available at http://www.nws.noaa.gov/directives/

There is no easy way to create a unit to compare health and money. This study will focus on giving relative comparisons.

## Data processing

First, we need to download and import the data

```{r dataImport}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","stormData.csv.bz2")
importedstormData <- read.csv("stormData.csv.bz2")
str(importedstormData)
```

The damage columns are not exploitable yet, because the exposant is in another column. We need to create columns with the final value, other values than B, M and K are not significant and can be ignored.

```{r exponentialValuesCleaning}
# preparing data to be cleaned
stormData <- importedstormData
expvalues <- cbind(c("[^BbMmKk]","B","M","K"),c(1,1e+09,1e+06,1e+03))
stormData$PROPDMGEXP <- as.character(stormData$PROPDMGEXP)
stormData$CROPDMGEXP <- as.character(stormData$CROPDMGEXP)

# cleaning any other operator than B M K and replacing it with 1
stormData$PROPDMGEXP <- gsub("[^BbMmKk]","1",stormData$PROPDMGEXP)
stormData$CROPDMGEXP <- gsub("[^BbMmKk]","1",stormData$CROPDMGEXP)

# replaceing the B M K with proper exposant operator
for (i in 1:length(expvalues[,1])){
        stormData$PROPDMGEXP <- gsub(expvalues[i,1],expvalues[i,2],stormData$PROPDMGEXP,ignore.case = TRUE)
        stormData$CROPDMGEXP <- gsub(expvalues[i,1],expvalues[i,2],stormData$CROPDMGEXP,ignore.case = TRUE)
}

# giving the numeric class and getting the final value
stormData$PROPDMG <- as.numeric(stormData$PROPDMG)
stormData$PROPDMGEXP <- as.numeric(stormData$PROPDMGEXP)
stormData$PROPDMGVALUE <- stormData$PROPDMG * stormData$PROPDMGEXP

stormData$CROPDMG <- as.numeric(stormData$CROPDMG)
stormData$CROPDMGEXP <- as.numeric(stormData$CROPDMGEXP)
stormData$CROPDMGVALUE <- stormData$CROPDMG * stormData$CROPDMGEXP
```

## Population health incidence

Let's find which events are the most harmful considering the population health. We will order our data with the "INJURIES" variable. We will then aggregate the result to avoid the doubles.

```{r harmfulInjuries}
stormDataHarmfulInjuries <- aggregate(INJURIES~EVTYPE,stormData,sum)
stormDataHarmfulInjuries <- stormDataHarmfulInjuries[order(-stormDataHarmfulInjuries$INJURIES),]
head(stormDataHarmfulInjuries)
```

That might be interesting to compare it to "FATALITIES"

```{r harmfulFatalities}
stormDataHarmfulFatalities <- aggregate(FATALITIES~EVTYPE,stormData,sum)
stormDataHarmfulFatalities <- stormDataHarmfulFatalities[order(-stormDataHarmfulFatalities$FATALITIES),]
head(stormDataHarmfulFatalities)
```

The most harmful event is the tornado, common for injuries and fatalities.

Let's compare the 5 most harmful fatalities and injuries events.
```{r fatalitiesInjuriesComparison}
topFatalities <- stormDataHarmfulFatalities[1:5,1]
topInjuries <- stormDataHarmfulInjuries[1:5,1]
topharmful <- c(as.character(topFatalities),as.character(topInjuries))
healthimpact <- join(stormDataHarmfulFatalities,stormDataHarmfulInjuries)
healthimpact$top <- healthimpact$EVTYPE %in% topharmful
healthimpact <- subset(healthimpact,healthimpact$top==TRUE)
healthimpact <- healthimpact[,1:3]
healthimpact <- melt(healthimpact, id="EVTYPE")
healthgraph <- ggplot(data=healthimpact,aes(EVTYPE,value)) + geom_point() + facet_grid(variable~.,scales="free")
healthgraph
```

The tornado event is significantely the most harmful, whereas the other events seems to be more similar.

## Economic incidence

There is 2 kind of economic values : property and crop. Using the processing we did, we can aggregate the datas
```{r economicConsequences}
# property consequences
stormDataEcoConsequencesproperty <- aggregate(PROPDMGVALUE~EVTYPE,stormData,sum)
stormDataHarmfulEcoConsequencesproperty <- stormDataEcoConsequencesproperty[order(-stormDataEcoConsequencesproperty$PROPDMGVALUE),]

# crop consequences
stormDataEcoConsequencescrop <- aggregate(CROPDMGVALUE~EVTYPE,stormData,sum)
stormDataHarmfulEcoConsequencescrop <- stormDataEcoConsequencescrop[order(-stormDataEcoConsequencescrop$CROPDMGVALUE),]

# merge datas
stormEcoconsequences <- join(stormDataHarmfulEcoConsequencesproperty,stormDataHarmfulEcoConsequencescrop)

# compute the total value
stormEcoconsequences$TOTALVALUE <- stormEcoconsequences$PROPDMGVALUE + stormEcoconsequences$CROPDMGVALUE
stormEcoconsequences <- stormEcoconsequences [order(-stormEcoconsequences$TOTALVALUE),]

# return the most economic incidences
head(stormDataHarmfulEcoConsequencescrop)
head(stormDataHarmfulEcoConsequencesproperty)
head(stormEcoconsequences)
```

When combining both crop and property damages, the most economic incidence event is the flood.

```{r ecograph}
ecograph <- ggplot(stormEcoconsequences[1:5,],aes(EVTYPE,TOTALVALUE)) + geom_point()
ecograph
```

We can see the flood is significantly more harmful than the other events.

## Results

For each kind of damage (health and economy), we observed that there is one particular event looking really more harmful than others.

The most harmful consequence for health is tornados, whereas the event with the most economic incidence is the flood.

Interesting fact : the most harmful event for health doesn't seem really harmful for economy, and conversely.