---
title: "Characteristics of major storms and weather events in the United States"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# ajouter le cache
# retirer les output des librairies chargées
library(dplyr)
```

## Abstract



## Data processing

First, we need to download and import the data

```{r dataImport}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","stormData.csv.bz2")
stormData <- read.csv("stormData.csv.bz2")
str(stormData)
```

## Population health incidence

Let's find which events are the most harmful considering the population health. We will order our data with the "INJURIES" variable. We will then aggregate the result to avoid the doubles.

```{r harmfulInjuries}
stormDataHarmfulInjuries <- aggregate(INJURIES~EVTYPE,stormData,sum)
stormDataHarmfulInjuries <- stormDataHarmfulInjuries[order(-stormDataHarmfulInjuries$INJURIES),]
head(stormDataHarmfulInjuries)
```

That might be interesting to compare it to "FATALITIES"

```{r harmfulFatalities}
stormDataHarmfulFatalities <- aggregate(FATALITIES~EVTYPE,stormData,sum)
stormDataHarmfulFatalities <- stormDataHarmfulFatalities[order(-stormDataHarmfulFatalities$FATALITIES),]
head(stormDataHarmfulFatalities)
```

## Economic incidence

```{r economicConsequences}
stormDataEcoConsequences <- aggregate(PROPDMG~EVTYPE,stormData,sum)
stormDataHarmfulEcoConsequences <- stormDataEcoConsequences[order(-stormDataEcoConsequences$PROPDMG),]
head(stormDataHarmfulEcoConsequences)
```

## Results